\chapter*{Abstract} 
\addcontentsline{toc}{chapter}{Abstract}
\pagestyle{plain}

The abstract should consist of a single paragraph (no paragraph breaks), and should not exceed 200 words. It should offer a compact overview of the thesis and help readers quickly assess its relevance. The abstract must cover the following five components:

\paragraph{Introduction}
Briefly present the research topic and context.

\textit{Example:} 
This thesis investigates how transformer-based models can be optimized for low-resource syntactic parsing, a crucial task for inclusive NLP development.

\paragraph{Research Question} 

State the primary research question clearly.

\textit{Example:} 
The main question addressed is: How can fine-tuning strategies be adapted to improve syntactic parsing performance in low-resource languages?

\paragraph{Method} 

Summarize the methodology used.

\textit{Example:} 
The study applies transfer learning using pre-trained multilingual models and evaluates performance across four low-resource language datasets.

\paragraph{Results} 

Outline the most important findings.

\textit{Example:} 
Results show that adapter-based fine-tuning yields competitive accuracy while reducing model size and training time significantly.

\paragraph{Discussion} 

Mention implications, limitations, or potential applications.

\textit{Example:} 
These findings suggest scalable solutions for syntactic parsing in underrepresented languages, with implications for equitable access to NLP tools.