\chapter{Introduction} \label{cha:introduction}
% why is this thesis important?
% what is this study to achieve?
% provide enough background to make work comprehensible to intended audience
% key components:
%   context and motivation
%   research problem without technical/methodological details
%   brief literature context
%   clarify overall goal and set expectations for the reader
%   research questions
%   define collaboration if the thesis is conducted in collaboration
%   provide a brief overview of the structure of the thesis. describe what each chapter covers.
% 1-1.5pages
\pagestyle{plain}

Large language models almost universally process text as sequences of discrete subtokens rather than whole words \citep{mielke2021}, which can split meaningful units into smaller fragments. In the GPT-2 tokenizer, for example, the name `Hemingway' is represented as \texttt{['H',~'eming',~'way']}. This raises a question regarding how models recover larger semantic units from these fragments, given they show robustness to corruption \citep{cao2023}. One hypothesis, often referred to as \textit{detokenization}, suggests that early layers in the model partially reverse the tokenization process by combining subtokens into more coherent segments. Evidence for this idea stems from interpretability work such as \citet{elhage2022}, which found neurons in early layers that respond strongly to token boundary artifacts. Similar hierarchical patterns appear in computer vision \citep{jastrzebski2018}, with for example \citet{zeiler2014} demonstrating early convolutional layers detecting simple features like edges while deeper layers capture more complex structures.

The mechanism of detokenization is described by \citet{kaplan2025} as constituting (at least) two components inside LLMs: \textit{attention heads/mass} which are active in the very early layers, and the \textit{feedforward network} in the early-mid layers which is responsible for refining those initial attended subtokens into the final token representation. Their hypothesis is that modern decoder LLMs possess an \textit{inner lexicon} which goes beyond the limits of the tokenizer, and that the detokenization process is effectively the model's attempt to map the input embeddings to that internal lexicon for processing in later inference stages.

This thesis examines whether detokenization depends on specific attention heads across languages (Chinese, Turkish, and English) and how the process varies under different tokenization schemes. In particular, we compare BPE- and SentencePiece-based tokenizers with byte-based tokenizers for otherwise identically trained models. We also investigate how this phenomena is affected by parameter size by analyzing a range of parameter sizes for one model family. Using activation patching, specifically PatchScopes \citep{ghandeharioun2024}, we probe word retrieval across layers from both original and attention head ablated inputs. This allows us to test whether specific attention heads are critical for detokenization or whether later components can compensate for earlier errors. Overall, we aim to improve the empirical understanding of detokenization by examining the effects of tokenizer design and multilingual input.

Section \ref{sec:background} gives an overview of the various tokenization schemes, the Transformer architecture and self-attention mechanism, and the field of mechanistic interpretability. Section \ref{sec:methodology} presents the data, models, and methodology that the research questions

\section{Research Questions}

\begin{description}
\item[\textbf{RQ1}] Do a small set of specific attention heads play a crucial role in forming the intermediary representation of whole words?
\item[\textbf{RQ2}] Does the choice of tokenizer play a role in how well the intermediary representation is formed and how robust the model is to attention head ablation?
\item[\textbf{RQ3}] Does the parameter size play a role in how well the intermediary representation is formed and how robust the model is to attention head ablation?
\end{description}

\section{Collaboration}

The thesis is done in collaboration with the NLP group at RISE.
