\chapter{Background} \label{cha:background}
% aim for 5-7 pages
% suggested structure:
%     topic introduction
%     key concepts and definitions
%     literature review
%         foundational landmark studies
%         recent developments and trends
%         gaps, limitations, unresolved challenges
%     position thesis in relation to literature
%         research questions

% introduce tokenization
% introduce autoregressive language models
% introduce mechanistic interpretability as a field
%   detokenization

Large language models (LLMs) do not process text as sequences of words. Instead, their input is segmented into tokens, many of which correspond only to fragments of words. For example, the city name \textit{Paterson} is represented in the GPT-2 tokenizer as \texttt{['Pat', 'erson']}. Because all subsequent computation in the model operates on the basis of those original tokens, any information that spans multiple tokens must be reconstructed during inference. Understanding how models perform this reconstruction is therefore important for explaining their robustness to typos \citep{zhao2026evaluatingrobustnesslargelanguage, cao2023} and non-canonical segmentation \citep{gutierrez2023, zheng2026} This requires understanding how tokenization shapes model input, how this input is transformed within the model, and how prior work has analyzed internal computations in LLMs.

This chapter is structured as follows: Section~\ref{sec:background_tokenization} introduces common tokenization algorithms, Section~\ref{sec:background_transformer} outlines the relevant components of the transformer architecture, and Section~\ref{sec:background_interpretability} presents work in mechanistic interpretability, including the \textit{detokenization hypothesis}, which proposes that early layers reconstruct larger linguistic units from fragmented inputs.

\section{Tokenization in LLMs}\label{sec:background_tokenization}

Compression-based subword tokenization algorithms are the standard for modern LLMs \citep[Chapter 2]{jm3}. The primary motivation for subword tokenization is to mitigate the out-of-vocabulary problems of earlier word-based approaches \citep{mielke2021}. Despite their arbitrary nature, tokens have been shown to reflect semantics \citep{haslett2025semantics} and psychological saliency \citep{haslett2025psychological}. In practice, several related approaches are used to construct such subword vocabularies, differing primarily in how they segment text and assign vocabulary candidates. The most widely used methods include \textbf{Byte Pair Encoding (BPE)}, \textbf{Unigram} language model tokenization, and \textbf{byte-level} tokenization. These approaches represent different trade-offs.

\begin{description}

\item[Byte-Pair Encoding]
Byte-Pair Encoding (BPE), introduced by \citet{bpe2} and adapted for neural machine translation by \citet{bpe}, constructs a subword vocabulary by iteratively merging frequent symbol/byte pairs. In practice, BPE tokenization involves a \textit{pre-tokenization} step followed by a merge learning procedure. Pre-tokenization defines the initial units over which merges are applied, often using simple heuristics such as splitting on whitespace and punctuation \citep{schmidt2025}. For example, the GPT-2 tokenizer \citep{radford2019} uses a regular-expression-based pre-tokenization scheme with a byte-level fallback. Given the input text \textit{``It's always 5 O'clock in New Orleans''} GPT-2 pre-tokenization produces \texttt{['It', "'s", 'Ġalways', 'Ġ5', 'Ġo', "'", 'clock', 'Ġin', 'ĠNew', 'ĠOrleans']}, where \texttt{'Ġ'} refers to a unique representation for whitespace. These opinionated pre-tokenization decisions can lead to unintuitive segmentations, particularly in languages which don't adhere to those design decisions. Some newer approaches, like \citet{liu2026superbpe}, propose using a two-fold learning process to learn tokens \textit{across} token boundaries, leading to better compression. However, such approaches are not widespread. For insufficiently trained byte-based BPE tokenizers, we can end up with \textit{glitch} tokens due to how languages such as Chinese is represented in bytes, where each character is $\sim$ 3 bytes, and where many of those bytes don't correspond to anything meaningful by themselves. \citet{jang2025} show that these can introduce a potential vulnerability when they occur next to unfamiliar tokens.

\item[Unigram]
Unigram tokenization \citep{kudo2018} models tokenization as a probabilistic process, selecting segmentations based on a learned distribution over possible subword units using a Unigram model. Unlike BPE, it does not rely on deterministic merging rules and can represent multiple valid segmentations. This flexibility is particularly useful in multilingual settings, where assumptions about token boundaries (e.g., whitespace) may not hold. Implementations such as SentencePiece \citep{sentencepiece} avoid explicit pre-tokenization, enabling more language-agnostic processing by not making assumptions what a token \textit{can} be beforehand.

\item[Byte-level Tokenization]
Some approaches, such as ByT5 \citep{xue2022}, operate directly on raw byte sequences without applying compression. This reduces preprocessing assumptions and can improve robustness to unseen inputs in some cases \citep{aleph2026}, but results in longer input sequences and more fragmented representations. The token mapping is trivial. For example, the \texttt{By-T5} tokenizer can be replaced wholly by one line of Python code. Let \texttt{text} be a string you want tokenized. 3 denotes the number of reserved special tokens:
\begin{equation}
    \texttt{[int(x) + 3 for x in text.encode("utf-8")]}
\end{equation}

\end{description}

\citet{arnett2025, wegmann2025} show that pre-tokenization choices can disproportionately affect non-Latin languages, leading to suboptimal segmentations and degraded performance in multilingual settings. More generally, tokenization and model design in LLMs are often English-centric \citep{joshi2020, blasi2022}, resulting in lower performance and higher computational costs for other languages \citep{ahia2023, petrov2023}.

\section{Transformer Architecture}\label{sec:background_transformer}

Most modern large language models are autoregressive decoder-only models based on the transformer architecture introduced by \citet{vaswani2017}. At a high level, a decoder-only transformer processes a sequence of token representations and builds contextualized embeddings through a series of layered transformations. In each layer, token representations are gradually updated by incorporating information from other tokens in the sequence. The resulting representations are then used to predict the next token in the sequence.

Before these transformations can take place, the discrete tokens produced by the tokenizer must be converted into vector representations that the model can process. This is done using an input embedding layer, which maps each token in the vocabulary to a dense vector. Since the transformer architecture does not inherently encode sequence order, positional information is added to these embeddings to allow the model to distinguish between different token positions in the input. Modern LLMs, like Llama \citep{llama2} and Qwen \citep{qwen2, qwen2.5} tend to utilize RoPE \citep{rope} positional encodings where instead of adding position embeddings we \textit{rotate} its input by an angle. This results in the model encoding \textit{relative} position differences rather than absolute position differences. The resulting representations (token embeddings + positional encodings) are then passed through a stack of transformer blocks, which iteratively refine the representations. After the final layer, a normalization step and a linear projection map the representations to the vocabulary space, producing a probability distribution over possible next tokens. This high-level overview is pictured in Figure~\ref{fig:transformer_layers}.

\input{fig/transformer_layers.tex}
\input{fig/transformer_block.tex}

Transformer blocks consist of a self-attention layer and a feed-forward network, each wrapped with residual connections and layer normalization, visualized in Figure~\ref{fig:transformer_block}. The central concept in the transformer block is the self-attention layer. In a self-attention layer, three linear projections of an input sequence with dimensionality $d$ produces the \textit{Query}, \textit{Key}, and \textit{Value} vectors. These vectors $Q$, $K$, and $V$ are then used to calculate attention scores:

\begin{equation}\label{eq:attention}
\text{Att}(Q, K, V) = \text{softmax} \left( \frac{QK^{\top}}{\sqrt{d}} \right) V
\end{equation}

\noindent
with softmax being defined as

\begin{equation}\label{eq:softmax}
\text{softmax}(z_i) = \frac{e^{z_i}}{\sum_{j=1}^{k} e^{z_j}}
\end{equation}

\noindent
where $z = (z_1, \dots, z_k)$ is a vector of input scores, and as output gives a normalized probability distribution from every token to every token summing to $1$. As noted by \citet{correia2019} this normalization makes it possible, in principle, for tokens to incorporate information from all other tokens in the sequence. In decoder-only transformers, training is performed using next-token prediction, where attention scores for tokens occurring after the current token are masked. To allow the model to capture different types of relationships in parallel, transformers use \textit{multi-head attention}. This involves applying Equation \ref{eq:attention} many times in parallel, forming several \textit{attention heads} which can attend to distinct parts of the sequence. Each head is calculated as such, where $W_i$ represents the learnable weight matrices for the $i$-th attention head:

\begin{equation}\label{eq:multihead}
\text{head}_i (Q, K, V) = \text{Att}(QW_i^{Q}, KW_i^{K}, VW_i^{V})
\end{equation}

Many modern LLM families, such as Llama2 \citep{llama2} and Qwen \citep{qwen2.5,qwen2}, utilize Grouped Query Attention \citep{ainslie2023}, where multiple heads share keys and values. This leads to faster training and inference, at the potential cost of increased interactions between heads compared to standard multi-head attention.
% Also worth noting is the layer normalization, whose task is to help stabilize training, which differs depending on the implementation. Llama2 and Qwen use RMSNorm \citep{rmsnorm}, which unlike LayerNorm \citep{layernorm} doesn't perform mean-centering, potentially destabilizing training for smaller models.

\section{Mechanistic Interpretability}\label{sec:background_interpretability}

While the mathematical operations and components of neural networks, including transformers, are well understood, explaining how specific outputs emerge remains a challenge. This has motivated the field of mechanistic interpretability, which aims to understand how model behavior emerges by reverse engineering the computations. Early efforts were mostly concerned with vision \citep{cammarata2020a}, and focused on identifying visually interpretable features. For example, features were found in neural models corresponding to grocery store detectors \citep{nguyen2016}, curve detectors \citep{cammarata2020b}, or even multi-modal features \citep{goh2021}.  However, extending these approaches to language directly is challenging as the representations are less directly interpretable\footnote{\citet{elhage2022a, templeton2024, bricken2023} argues this is because neurons in LLMs effectively are in \textit{super-position}, meaning that one neuron's direction in feature space corresponds to many separate features. A detailed discussion of this phenomenon is beyond the scope of this thesis.}

One approach is to analyze model behavior in terms of interacting components, often described as \textit{circuits}, that contribute to specific outputs. As argued by \citet{elhage2021}, all components of the transformer can be viewed as participating in such circuits through their interactions via the \textit{residual stream}. The residual stream acts as a ``communication channel'': each transformer block receives its input from it, and both the attention and feed-forward network write their output back into it. It is shown in red in Figure~\ref{fig:transformer_block}. The result is an additive process, where the model's representation is built up incrementally as it moves through the network.

Multiple components contribute to this gradual building of the representation, but the attention mechanism provides a particularly useful point of analysis. Due to the multi-head structure of transformers (see Section~\ref{sec:background_transformer}), individual attention heads can become highly specialized for distinct tasks, and prior work has shown that only a small subset of heads are crucial for model performance in certain downstream tasks, while many can be pruned with minimal impact \citep{voita2019}. \citet{liu2025_language_attention_heads} find a circuit of language-specific heads and show that these can guide generation toward specific languages. \citet{hanna2023} identify a circuit in GPT-2 that captures the ``XX greater than YY'' relation. \citet{carrasco2024} found a circuit of 8 attention heads involved in three-letter acronym detection. Even complex behaviors have been linked to a small number of attention heads; for example, \citet{zhou2025} show that ablating even a single head in LLaMA2-7B leads to a 16x increase in harmful responses. These findings suggest that attention heads act as functionally distinct components within the transformer.

Specialization does not appear to occur arbitrarily across layers. \citet{lad2024} argues that large language models go through \textit{stages of inference}, where each stage is responsible for a functionally distinct type of processing: early layers perform \textit{\textbf{detokenization}}, where subtokens merge into larger lexical units; middle layers perform \textit{feature engineering}, which constructs intermediate semantic features; \textit{prediction ensembling} in middle-late layers, where candidate next-token predictions emerge; and in the final layers \textit{residual sharpening}, which concentrates probability mass onto a small set of tokens.

Previous studies support this stage-wise view for LLMs. \citet{vig2019} observe that attention patterns in middle layers align more closely with syntactic dependency relations. \citet{ferrando2024similarity} identify circuits for language-specific subject–verb agreement localized to later layers. \citet{gurnee2024} show using linear probes that representation about geographical places (e.g.\ place names) emerges in early-middle layers and then plateaus.

Further evidence for this structured, layer-wise processing comes from studies of multilingual representations. \citet{conneau2020} show that multilingual models can align representations across languages with minimal supervision, suggesting the existence of a shared latent embedding space rather than purely language-specific processing. Subsequent work provides more direct evidence for stage-like behavior across layers. \citet{wendler2024} find that multilingual processing follows a three-phase structure: early layers encode language-specific information, middle layers shift toward a more language-agnostic (or English-centric) representation space, and later layers return to language-specific representations for output generation. Similar patterns have been reported in a range of studies \citep{tamo2026, zhong2026, zhao2024, dumas2025, basirat2025}.

This hierarchical organization is consistent with patterns observed in other domains. In computer vision, early convolutional layers detect simple features such as edges, while deeper layers capture increasingly complex structures \citep{zeiler2014, jastrzebski2018}. Similar hierarchical processing has also been hypothesized in the human brain \citep{wacongne2011}.

Evidence for detokenization as an early-stage process comes from studies of subtoken merging behavior. \citet{elhage2021} demonstrate in toy transformer models that attention heads can implement skip-trigram-like mappings, effectively reconstructing token sequences from fragmented inputs. For example, a tokenized sequence of ``Cindy'' without a space \texttt{["C", "indy"]} can be mapped back to the more typical form \texttt{["␣Cindy"]} through specific attention head patterns. Related work has also identified neurons in early layers that respond strongly to token boundary artifacts \citep{elhage2022a}, suggesting that models encode information about subtoken structure at early stages. Similar patterns have been identified in larger models. \citet{ferrando2024informationflow} construct information flow graphs and identify heads that preferentially attend within word boundaries, which they interpret as subword-merging heads. Similarly, \citet{lad2024} find attention heads that exhibit consistent patterns of subtoken merging, where later subtokens attend to earlier subtokens within the same word. \citet{kamoda2025} identify attention heads involved in detokenization through weight-based analysis of GPT-2 models, i.e.\ without forward passes.

\citet{kaplan2025} argue that detokenization is a two-stage process involving both attention mechanisms and feed-forward networks, and that it should be understood not merely as subword merging, but as a mapping from fragmented token inputs to an \textit{implicit/internal vocabulary}. This view builds on work by \citet{feucht2024}, who train linear probes across layers and show that representations of multi-token entities, words, and phrases undergo a form of ``erasure,'' where token-level information is concentrated in early layers. In this framework, attention mechanisms in early layers identify relevant tokens for constructing a coherent representation, while feed-forward networks in subsequent layers map this information into the model's internal vocabulary, building on prior work which shows feed-forward networks can act as key–value memory stores \citep{geva2021, geva2022, meng2022}.

\citet{kaplan2025} provide empirical support for this account by showing that attention weights are systematically higher within multi-token words than across word boundaries, and that ablating feed-forward outputs leads to greater degradation in performance than ablating the hidden layer output. These findings are consistent with a two-stage process in which attention identifies relevant subtoken information, and feed-forward networks integrate this information into coherent internal representations.
