\chapter{Results} \label{cha:results}

This section is structured as follows: first the main findings are summarized, grouped by research question. Then, in Section~\ref{sec:results_rq1} the results for \textbf{RQ1} is presented, followed by Section~\ref{sec:results_rq2} for \textbf{RQ2} and Section~\ref{sec:results_rq3} for \textbf{RQ3} respectively.

\begin{description}

\item[\textbf{RQ1}] Across both conditions, targeted ablation reduced retrieval performance more than random ablation for nearly all models, particularly at ablation levels $\geq$0.03. The largest gaps between targeted and random ablation were observed in the stronger-performing models and language settings.

\item[\textbf{RQ2}] Retrieval performance and targeted-ablation effects varied across tokenizer variants and languages. In English, \texttt{xglm} achieved the highest clean retrieval score (0.81) and exhibited larger targeted ablation effects than the other tokenizer variants, whereas the differences between tokenizer variants were smaller in Chinese and Turkish.

\item[\textbf{RQ3}] Larger models get higher retrieval performance and achieve a qualitatively notable early retrieval peak compared to smaller models which show more distributed retrieval patterns. Within the \texttt{qwen2.5} model family, retrieval performance generally decreased as targeted ablation strength increased from 1\% to 5\%.

\end{description}

\section{Targeted Ablation}\label{sec:results_rq1}

Targeted ablation leads to worse or identical retrieval than the random ablation in every case except two, where retrieval increases from 0.01 to 0.02 in Turkish for the \texttt{byt5} model under 5\% ablation, and the \texttt{aya} model in English under 1\% where it increases from 0.28 to 0.32 after targeted ablation. The largest drop occurs for the \texttt{qwen2.5} 7B model in English. Under 5\% ablation, targeted ablation drops retrieval from 0.79 to 0.01, while random drops it to 0.81. For 3\% ablation retrieval drops to 0.08 after targeted ablation, as compared to 0.73 under random ablation. With the exception of the \texttt{byt5}/Turkish and \texttt{aya}/English example already presented, none of the models are able to fully recover retrieval at a later stage after targeted ablation, though the models compensate at different degrees (see Figure~\ref{fig:ablation_stacked}). The final layer results for the Toksuite models is presented in Table~\ref{tab:toksuite_summary}, \texttt{qwen2.5} models in Table~\ref{tab:qwen_summary}.

\input{tab/toksuite_summary.tex}
\input{tab/qwen_summary.tex}

\section{Toksuite --- Tokenizer \& Language}\label{sec:results_rq2}

For the Toksuite models, the magnitude of targeted ablation depends by the language and tokenizer. The strongest effect is observed for English with the \texttt{xglm} tokenizer, where cumulative retrieval drops from $\sim$0.7 to $\sim$0.15 for the first 25\% of layers under 5\% ablation (shown in Figure~\ref{fig:toksuite_stacked}). Although performance partially recovers in later layers, it remains 0.38 below the clean baseline at the final layer. In contrast, targeted ablation has minimal effect in Chinese/turkish for certain tokenizers. For the \texttt{byt5} tokenizer, no additional degradation is observed under targeted ablation, and in one case leading to \textit{improvements}. Similarly, the \texttt{gpt2} tokenizer shows only a small drop (0.05/0.04) in Chinese/Turkish under 5\% ablation, less than other tokenizer/language combinations.

Averaged across Toksuite models, retrieval performance is highest for English, lower for Turkish, and lowest for Chinese. In Chinese, most tokenizers perform similarly, with the exception of \texttt{byt5}, which deviates from this trend. Figure~\ref{fig:toksuite_stacked} presents the cumulative retrieval for each tokenizer under each ablation level, pooled across languages. Figure~\ref{fig:toksuite_rr} shows the retrieval rate per layer for each tokenizer and condition.

\input{fig/rr.tex}
\input{fig/stacked.tex}

\section{\texttt{qwen2.5} --- Parameter Size}\label{sec:results_rq3}

The \texttt{qwen2.5} models exhibit a qualitatively different pattern. Here, targeted ablation has a particularly big impact on the largest 7B model for 3/5\% ablation. For example, retrieval drops from 0.79 to 0.08/0.01 under targeted ablation in English under 3/5\% ablation. Random ablation has a larger effect in these models than in Toksuite, but remains weaker than targeted ablation. The highest random ablation impact is for 5\% ablation 7B/English, where random ablation reduces retrieval by 0.18 as compared to a 0.78 reduction under targeted ablation. Across all \texttt{qwen2.5} models, English consistently outperforms Turkish and Chinese, which show similar performance levels. The largest model has the best retrieval of the tested sizes, although the 1.5B outperforms the 3B in all languages except Turkish.

Layerwise analysis reveals that the 7B and 3B models exhibit early peaks in retrieval rate, while the 1.5B model distributes retrieval more evenly across layers. The pooled layerwise analysis is available in Figure~\ref{fig:qwen_stacked}. The 0.5B model shows neither strong retrieval nor a clear peak. Retrieval rate curves are shown in Figure~\ref{fig:qwen_rr}.
