% discussion should:
% answer the research questions based on results
% reflect on strengths/weaknesses
% show how findings relate to the background and literature
% suggested structure:
% summary of main results: focus on patterns, contrasts, significant observations
% interpretation/comparison: analyze what the results suggest; why certain models performed the way they did, surprising or unexpected outcomes, how findings relate to previous studies
% methodological reflections
% limitations
% applications and implications for the field
% future research
\chapter{Discussion} \label{cha:discussion}

This chapter begins with a summary of the main findings in relation to the research questions. Section~\ref{sec:discussion_tokenizer} discusses the effects of tokenizer choice and Section~\ref{sec:discussion_qwen} on model size, followed by a reflection on limitations and implications in Section~\ref{sec:discussion_limitations}, and directions for future work in Section~\ref{sec:discussion_futurework}.

\begin{description}
\item[\textbf{RQ1}] The results show targeted ablation has a disproportionately large impact on word retrieval compared to random ablation. In most cases, models are unable to recover retrieval performance after targeted ablation, and the retrieval peak is consistently delayed. In extreme cases, such as the \texttt{Qwen2.5} 7B model, retrieval collapses from 0.79 to 0.01, while random ablation produces only minor changes. This pattern is consistent with findings by \citet{voita2019}, indicating that a small subset of attention heads plays a specialized and critical role.

\item[\textbf{RQ2}] Tokenizer choice has impact on the formation of intermediary representations. Differences in retrieval performance cannot be explained by compression efficiency alone. For example, the \texttt{XGLM} tokenizer achieves substantially higher retrieval than \texttt{gpt2} in English (0.81 vs.\ 0.47), despite \texttt{gpt2} being better at compressing English. This suggests that tokenizer design has an impact on how the representations are structured within the model during training. At the same time, robustness to ablation does not follow a consistent pattern across tokenizers, and varies depending on language.

\item[\textbf{RQ3}] Larger models achieve higher retrieval scores, with the 7B model performing best overall and the 0.5B model performing worst. Mid-sized models (1.5B and 3B) show comparable performance, with variations across languages. Larger models show different retrieve rate patterns, with a distinct peak at early layers which is not present in the 1.5B and 3B models, that instead show a more distributed retrieval pattern. This suggests that scaling leads to more efficient but less distributed representations as models get bigger.

\end{description}

\section{Tokenizer Impact}\label{sec:discussion_tokenizer}

The impact of tokenizer choice can be partly explained by differences in compression and the resulting training dataset size in \texttt{Toksuite} models. Because all models are trained on a fixed number of tokens, tokenizers with lower compression rates expose the model to less raw text. As shown in Table~\ref{tab:fertility}, \texttt{byt5} operates on 1 byte/token, meaning that it is effectively trained on less data than more compressive tokenizers such as \texttt{xglm} which gets $\sim$4 bytes/token. This reduced exposure may limit the model's ability to learn specialized attention patterns. Instead of relying on a small set of heads to reconstruct word-level representations, the model may develop more distributed mechanisms for combining subtoken information, reducing the impact of targeted ablation.

However, compression alone does not explain the observed differences. For example, the \texttt{gpt2} tokenizer achieves a compression rate close to \texttt{byt5} for Chinese (1.36 vs.\ 1.0), yet attains substantially higher retrieval performance (0.36 vs.\ 0.12). Similarly, the \texttt{xglm} tokenizer outperforms \texttt{gpt2} in English despite slightly worse compression. These comparisons indicate that factors beyond compression and effective training signal must play a role.

One possible contributing factor is cross-lingual transfer in the \texttt{gpt2} / Chinese example. Prior work using BERT models has shown that monolingual models can learn abstractions that generalize across languages \citep{artetxe2020}, and that embedding alignment correlates with transfer performance \citep{deshpande2022}. In this context, the \texttt{gpt2}-based model may benefit from patterns learned from English and Turkish when processing Chinese, allowing it to better organize fragmented token sequences. Though, we see explicitly multilingual tokenizers like \texttt{aya} tokenizer gets \textit{worse} retrieval than the \texttt{gpt2} in both English and Chinese, despite being trained on more data in virtue of its more efficient tokenizer, suggesting that there are interactionary cross-lingual effects (i.e.\ the \textit{curse of multilinguality}) which can harm downstream performance for all languages in multilingual models \citep{chang2024}.

Another explanation could be that the efficiency of a tokenizer influences how representations are distributed within the model. When tokenization is poor, the model may encode word-level structure across a broader set of components, resulting in more distributed representations since it has to compile more information. In contrast, better tokenization may allow the model to concentrate this functionality in a smaller number of specialized attention heads. This distinction provides a possible explanation for the observed differences in robustness: models with more distributed representations are less sensitive to targeted ablation, while models relying on specialized components degrade more sharply when those components are removed. This interpretation is supported by attention heatmaps for Chinese (Figure~\ref{fig:gpt2_heatmaps}), where our attention patterns are qualitatively different than in English and Turkish, suggesting weaker specialization. Additionally, byte-level artifacts such as the ``ghost token'' effects described by \citet{jang2025} may further disrupt the formation of coherent internal representations, particularly in languages like Chinese and Turkish where such byte artifacts would occur more frequently for poorly trained tokenizers. This is consistent with findings from previous studies that show the importance of pre-tokenization choices \citep{wegmann2025, arnett2025}.

\input{fig/gpt2_heatmap.tex}

Beyond compression and representational distribution, tokenizer architecture itself may also contribute to the observed differences. The \texttt{xglm} tokenizer achieves the highest retrieval performance in English, outperforming even larger models such as \texttt{qwen2.5-7B}, despite having slightly worse compression than \texttt{gpt2}. This suggests that differences in tokenization strategy, such as the use of a Unigram language model versus BPE, may influence how effectively embeddings and intermediate representations are learned. While prior work (e.g.\ \citet{wu2023}) emphasizes the disproportional impact of the embedding layer compared to the tokenizer, the results indicate that tokenizer design can shape how efficiently embeddings are trained. However, given that only one Unigram-based model is included, this remains speculative and warrants further investigation.

\section{Parameter Size Impact}\label{sec:discussion_qwen}

Retrieval performance generally increases with parameter size, with the 0.5B model performing worst and the 7B model performing best within the \texttt{qwen2.5} family. This is consistent with prior findings that embeddings of high-frequency words converge quickly during training, while low-frequency words take longer and retain alignment with the bias in their random initializations.

However, this relationship does not strictly hold. The 1.5B model slightly outperforms the 3B model in English and Chinese, despite similar overall retrieval levels. This suggests that increasing parameter size does not uniformly improve representation quality across configurations. One interpretation is that scaling changes \textit{how} representations are constructed within the model. Smaller models appear to rely on more distributed mechanisms, while larger models develop specialized components.

This shift is reflected in the layerwise retrieval patterns (Figure~\ref{fig:qwen_rr}). The 7B and 3B models exhibit distinct early peaks (at less than 25\% of model depth), followed by a rapid decline, indicating that key representations are formed early and handled by specific components. In contrast, the 0.5B and 1.5B models show more gradual and distributed accumulation of retrieval across layers. Similar patterns have been observed in larger models \citep{kaplan2025}, as well as in studies showing that structured behaviors such as induction are less clearly localized in smaller models \citep{feucht2025}.

Within this framework, the 1.5B model may represent an intermediate regime, where layer specialization is emerging but not yet fully established. This could explain why it performs competitively with the 3B model in some cases, while still exhibiting more distributed behavior across layers. Despite this apparent specialization, no clear distinction in robustness to ablation is observed between the different sizes. One possible explanation is that the ablation setup may not be fine-grained enough to capture differences in sensitivity, as removing a relatively large number of heads could mask more subtle variations. This remains speculative and would require more targeted ablation strategies to verify.

\section{Implications \& Future Work}\label{sec:discussion_futurework}

The results suggest that representation structure in transformer models is shaped not only by architecture and scale, but also by tokenizer design, with consequences for both research and practice. The degree of functional specialization appears to vary with tokenization, implying that interpretability findings based on localized components may not generalize across model configurations. For model development, the results suggest tokenizer choice influences not just the efficiency of learning embeddings, but influences how information is distributed within the model.

There are several paths of future work. This study does not consider the impact ablation has on generation. It could be the case that ablation does not hinder the model's ability to generate good next-token predictions. One potential future work could take inspiration from \citet{basile2026} who managed to \textit{steer} output using semantic attention head ablation, but redesign the task for detokenization. Another unexplored area is how the phenomena emerges during training. \citet{korner2026} studied EuroLLM checkpoints to see how shared concept spaces emerged during training. This is unexplored for the area of detokenization.

Several directions for future work remain. This study does not evaluate the effect of ablation on generation, and it is possible that strong degradation in retrieval does not translate to impaired next-token predictions. Future work could extend this by evaluating generation quality under targeted ablation, or by adapting approaches such as \citet{basile2026}, who demonstrate controlled output steering via attention head ablation, to detokenization tasks. Another open question is how these representational patterns emerge during training. Prior work by \citet{korner2026} analyzes the development of shared concept spaces across EuroLLM checkpoints; a similar approach could be applied to study the emergence of detokenization-related attention patterns.

\section{Limitations}\label{sec:discussion_limitations}

Ablation is an inherently interventional method and may induce non-typical model states. As noted by \citet{hase2021}, observed failures under ablation may reflect artifacts of the intervention rather than naturally occurring behavior. While this study focuses on internal representations rather than downstream outputs, the possibility remains that the measured effects do not fully correspond to how the model operates under normal conditions.

Another limitation concerns the interpretation of attention head importance. As argued by \citet{wiegreffe2019}, the apparent importance of specific components may be explained by correlated or redundant structures elsewhere in the model. Consequently, the results do not establish a strictly causal role for individual attention heads, but rather indicate their importance within the constraints of the ablation setup.

Finally, the ablation procedure operates at a relatively coarse granularity (5\%). This may obscure more fine-grained differences in sensitivity between models, particularly when comparing across scales or tokenization strategies.
