\chapter*{Appendix 3: Use of AI Tools in Thesis Production}

In this appendix, you must clearly state \textbf{if and how you used ChatGPT or any similar AI-based tools} (e.g., Copilot, Gemini, Claude, DeepL, Grammarly) during the development of your thesis. This applies whether the use was minimal, extensive, or non-existent.

\vspace{1em}

Please include the following information:

\begin{enumerate}
  \item \textbf{Tools Used} \\
  List the AI-based tools used during the research, writing, analysis, or editing process. Include:
  \begin{itemize}
    \item Tool name (e.g., ChatGPT, Grammarly, DeepL)
    \item Version or date of use (if applicable)
  \end{itemize}
  
  \item \textbf{Purpose of Use} \\
  For each tool, describe its purpose. Examples include:
  \begin{itemize}
    \item Brainstorming or refining research questions
    \item Generating or checking code
    \item Summarizing literature
    \item Improving language or grammar
    \item Translating text
    \item Drafting or revising sections
    \item Formatting references
  \end{itemize}
  
  \item \textbf{Scope and Extent} \\
  Indicate whether the use was:
  \begin{itemize}
    \item Minor (e.g., only grammar suggestions or occasional prompts)
    \item Moderate (e.g., regular use for code or summaries)
    \item Substantial (e.g., frequent use for drafting or structuring text)
  \end{itemize}
  
  \item \textbf{Critical Evaluation} \\
  Briefly explain how you ensured the accuracy, originality, and academic integrity of any content influenced by AI tools. For example:
  \begin{itemize}
    \item Cross-checking outputs with primary sources
    \item Revising or rewriting AI-generated text
    \item Ensuring compliance with ethical and academic standards
  \end{itemize}
\end{enumerate}

\vspace{1.5em}
\noindent\textbf{Example Statement:}
\begin{quote}
\textit{Tools Used: ChatGPT (GPT-4, March--May 2025), Grammarly. \\
Purpose of Use: We/I used ChatGPT to explore alternative phrasing for the methodology section and to get feedback on the logical flow of my/our arguments. Grammarly was used to correct grammar and spelling. \\
Scope and Extent: Moderate. \\
Critical Evaluation: All AI-generated outputs were critically reviewed and edited. No content was used without verification or adaptation. \\
}
\end{quote}